\section{Published Papers}

\begin{itemize}
    \item Petr Lozov, Ekaterina Verbitskaia, and Dmitry Boulytchev. \emph{``Relational interpreters for search problems.''} Relational Programming Workshop. 2019. \href{https://dash.harvard.edu/bitstream/handle/1/41307116/tr-02-19.pdf?sequence=1&isAllowed=y#page=47}{The full text is available here}.
    \begin{itemize}
        \item This paper formalizes the verifier-to-solver approach and highlights its pitfalls. 
    \end{itemize}
    \item Ekaterina Verbitskaia, Daniil Berezun, Dmitry Boulytchev. \emph{``An Empirical Study of Partial Deduction for \textsc{miniKanren}.''} Verification and Program Transformation Workshop. 2021. \href{https://cgi.cse.unsw.edu.au/~eptcs/paper.cgi?VPT2021.5.pdf}{The full text is available here}. 
    \begin{itemize}
        \item This paper describes a novel partial evaluation approach developed for \textsc{miniKanren}. 
    \end{itemize}
    \item Ekaterina Verbitskaia, Igor Engel, Daniil Berezun. \emph{``A Case Study in Functional Conversion and Mode Inference in \textsc{miniKanren}.''}  Partial Evaluation and Program Manipulation Workshop. 2024. \href{https://web.archive.org/web/20240111175807id_/https://dl.acm.org/doi/pdf/10.1145/3635800.3636966}{The full text is available here}.
    \begin{itemize}
        \item This paper describes functional conversion that makes it possible to generate a functional implementation of a solver from the functional implementation of a verifier, thus getting rid of the inefficiencies innate to the relational programming itself.  
    \end{itemize}

% http://minikanren.org/workshop/2020/minikanren-2020-paper2.pdf
% 
\end{itemize}


