\documentclass[crop=false]{standalone}
\usepackage[subpreambles=false]{standalone}
\usepackage{import}
\usepackage{blindtext}

\begin{document}

\begin{tabular}{| l | l |} 
  \hline
  Family Name, Given/First Name & Verbitskaia, Ekaterina \\ 
  \hline
  Matriculation number & 20333664 \\ 
  \hline
  What kind of thesis are you submitting & PhD‐Thesis \\
  \hline
 \end{tabular}

\vfill

\section*{English: Declaration of Authorship}

I hereby declare that the thesis submitted was created and written solely by myself
without any external support. Any sources, direct or indirect, are marked as such. I
am aware of the fact that the contents of the thesis in digital form may be revised with
regard to usage of unauthorized aid as well as whether the whole or parts of it may
be identified as plagiarism. I do agree my work to be entered into a database for it to
be compared with existing sources, where it will remain in order to enable further
comparisons with future theses. This does not grant any rights of reproduction and
usage, however.
The Thesis has been written independently and has not been submitted at any other
university for the conferral of a PhD degree; neither has the thesis been previously
published infull.


\section*{German: Erkl\"{a}rung der Autorenschaft (Urheberschaft)}

Ich erkl\"{a}re hiermit, dass die vorliegende Arbeit ohne fremde Hilfe ausschließlich von
mir erstellt und geschrieben worden ist. Jedwede verwendeten Quellen, direkter oder
indirekter Art, sind als solche kenntlich gemacht worden. Mir ist die Tatsache
bewusst, dass der Inhalt der Thesis in digitaler Form gepr\"{u}ft werden kann im Hinblick
darauf, ob es sich ganz oder in Teilen um ein Plagiat handelt. Ich bin damit
einverstanden, dass meine Arbeit in einer Datenbank eingegeben werden kann, um
mit bereits bestehenden Quellen verglichen zu werden und dort auch verbleibt, um
mit zuk\"{u}nftigen Arbeiten verglichen werden zu k\"{o}nnen. Dies berechtigt jedoch nicht
zur Verwendung oder Vervielf\"{a}ltigung.
Diese Arbeit wurde in der vorliegenden Form weder einer anderen Pr\"{u}fungsbeh\"{o}rde
vorgelegt noch wurde das Gesamtdokument bisher ver\"{o}ffentlicht.

\vfill

\hbox to \textwidth{\leaders\hbox to 5pt{\hss . \hss}\hfil}

\noindent
Date, Signature

\end{document}