\section{Conclusion}

In this paper we discussed some issues which arise in the area of partial deduction techniques for relational programming language \mk.
We presented a novel approach to partial deduction --- conservative partial deduction --- which uses a heuristic to select a suitable relation call to unfold at each step of driving.
We compared this approach with the most sophisticated implementation of conjunctive partial deduction --- \ecce partial deduction system --- on 6 relations which solve 2 different problems.

Our specializer improved the execution time of all queries.
\ecce worsened the performance of all 4 implementations of the propositional evaluator relation, while improving the other queries.
Conservative partial deduction is more stable with regards to the order of relation calls than \ecce which is demonstrated by the similar performance of all 4 implementations of the evaluator of logic formulas.

Some queries to the same relation were improved more by ConsPD, while others --- by \ecce.
We conclude that there is still not one good technique which definitively speeds up every relational program.
More research is needed to develop models capable of predicting the performance of a relation which can be used in specialization of \mk.
There are some papers which estimate the efficiency of partial evaluation in the context of logic and functional logic programming languages~\cite{vidal2004cost,vidal2008trace}, and may facilitate achieving this goal.
Employing a combination of offline and online transformations as done in~\cite{hybrid} may also be the step towards more effective and predictable partial evaluation.
Other directions for future research include exploring how specialization influences the execution order of a \mk program, improving ConsPD so that it succeeds at deforestation and tupling more often, and coming up with a larger, more impressive, set of benchmarks.